Higher order unification for free!

### 2024-04-09 paper/main.tex
some rework

The example at the beginning of \cref{sec:eta}, once
$\sigma = \{~ A \mapsto f ~\}$, triggers this rule since the link
becomes \linketaM{}{f}{\lambda x.B_{x}} and the \lhs is a constant.
In turn the rule runs $\lambda x.f \appsep x \Ue{} \lambda x.B_{x}$,
resulting in $\sigma = \{~ A \mapsto f ~;~ B_{x} \mapsto f ~\}$.
Decompilation the generates $\rho = \{~ X \mapsto f ~\}$, since
$X$ is mapped to $B$ and
$f$ is the $\eta$-contracted version of $\lambda x.f \appsep x$.

@@

% \noteE[inline]{IIRC, second order unification, where variables can stand for functions, is semi-decidable, indeed Huet's algorith enumerates all solutions. Third order, functions that take functions, is undecidable, proved by Dowek. IMO here the problem is that there is no unique solution, (semi)decidability is a secondary point. This paragraph needs fixing.}
\noteD[inline]{I've rewritten it, it is clearer?}
Until now, we have only dealt we unification of terms in \llambda. However, we
want the unification relation to be more robust so that it can work with terms
in \notllambda. In general, unification in \notllambda admits more then one
solution and committing one of them in the substitution does not guarantee
prop.~\ref{prop:complete}. For instance, $X \appsep a \Uo a$ is a unification
problem admits two different substitutions: $\rho_1 = \{X \mapsto \lambda x.x\}$
and $\rho_2 = \{X \mapsto \lambda \_.a\}$. Prefer one over the other may break
future unifications.

It is the case that, given a list of unification problems, $\foUnifPb_1\dots
\foUnifPb_n$ with $\foUnifPb_n$ in \notllambda, the resolution of
$\bigwedge_{i=0}^{n-1}\foUnifPb_i$ gives a partial substitution $\rho$, such
that $\rho\foUnifPb_n$ falls again in \llambda.
%
% make test ONLY=7003 TEX=TEX + deactivate beta tex
\printAlll
  {{{X,\lambda x.Y},
    {(X\appsep a),a}}}
  {{{A,\lambda x.B},
    {(A\appsep a),a}}}
  {{{Y,B,0},
    {X,A,0}}}
  {{}}

In the example above, we see that $\foUnifPb_1$ instantiates $X$ so that
$\foUnifPb_2$, can be solved in \llambda.
\noteE[inline]{it is even a ground term, there is no unification left to perform actually}
\noteD[inline]{i don't understand the note}
On the other hand, we see that, 
\Ue can't solve the compiled problems \hoUnifPb. In
fact, the resolution of $\hoUnifPb_1$ gives the substitution $\sigma = \{ A
\mapsto \lambda x. B\}$, but the dereferencing of $\hoUnifPb_2$ gives the 
non-unifiable problem $(\lambda x. B) \appsep a \Ue a$.

To address
this unification problem, term compilation should capture the
terms $t$ in \notllambda and replace them with fresh variables $X$.
The variables $X$ and the terms $t$ are linked through a \linkbeta.

@@

\newcommand{\rhsBetaHead}{\ensuremath{X_{s_1\dots s_n}}}
\newcommand{\rhsBeta}{\ensuremath{\rhsBetaHead\appsep t_1\dots t_m}\xspace}
% \noteE[inline]{Io direi $X_{s_1\ldots s_n}\appsep l_1\ldots l_m$
% ... where S = [s1 .. sn] e L = [l1 .. lm]}
\noteD[inline]{Is it clearer?}

\begin{invariant}[\linkbeta \rhs]
  The \rhs of any \linkbeta has the shape \rhsBeta such that $X$ is a
  unification variable with scope $s_1\dots s_n$ and $t_1\dots t_m$ is a list of
  terms. This is equivalent to \elpiIn{app[uva X S | L]}, where \elpiIn{S}~$=
  s_1\dots s_n$ and \elpiIn{L}~$=t_1\dots t_m$.
  \label{inv:beta-rhs}
\end{invariant}

@@

\begin{proof}[Proof sketch]
  Given $X \appsep t_1\dots t_n \Ue t$ where $t$ is a rigid term and
  $t_1\dots t_n$ is not in \llambda. By construction, $X \appsep t_1\dots t_n$
  is replaced with a variable $Y$, and the \linkbeta \linkbetaM{\Gamma}{Y}{X
  \appsep t_1\dots t_n} is created. The unification instantiates $Y$ to $t$,
  making the \lhs of the link a rigid term, while \rhs is still in \notllambda.
  The original problem is in fact outside \llambda.
\end{proof}

@@

Detection of \notllambda is quite simple to implement in the compiler, since it
is sufficient to detect applications with flexible head and argument that
are not in \llambda.

@@

\noteD{Paragraph for decompile in case of \maybebeta}
The \elpiIn{commit-link} predicate can be extended to add heuristics if during
the decompilation phase \linkbeta remain. For example, the same approximation
explained above can be delayed and applied only if the terms in \maybebeta never
falls in \llambda after the execution of all the unification problems. We want
to point out, that we call this \emph{approximation}, since we are
making a choice among all the possible unifiers and therefore, we can pick
the wrong one.

>>>>

The example at the beginning of \cref{sec:eta}, once
$\sigma = \{~ A \mapsto f ~\}$, triggers \progressetaleft since the link
becomes \linketaM{}{f}{\lambda x.B_{x}} and the \lhs is a constant.
In turn the rule runs $\lambda x.f \appsep x \Ue{} \lambda x.B_{x}$,
resulting in $\sigma = \{~ A \mapsto f ~;~ B_{x} \mapsto f ~\}$.
Decompilation the generates $\rho = \{~ X \mapsto f ~\}$, since
$X$ is mapped to $B$ and
$f$ is the $\eta$-contracted version of $\lambda x.f \appsep x$.

@@

% \noteE[inline]{IIRC, second order unification, where variables can stand for functions, is semi-decidable, indeed Huet's algorith enumerates all solutions. Third order, functions that take functions, is undecidable, proved by Dowek. IMO here the problem is that there is no unique solution, (semi)decidability is a secondary point. This paragraph needs fixing.}
% \noteD[inline]{I've rewritten it, it is clearer?}
In general, unification between \notllambda terms admits more then one
solution and committing one of them in the substitution does not guarantee
\cref{prop:complete}. For instance, $X \appsep a \Uo a$ 
admits two different substitutions: $\rho_1 = \{X \mapsto \lambda x.x\}$
and $\rho_2 = \{X \mapsto \lambda \_.a\}$. Prefer one over the other may break
future unifications.

It is the case, however, that, given a list of unification problems, $\foUnifPb_1\dots
\foUnifPb_n$ with $\foUnifPb_n$ in \notllambda, the resolution of
$\bigwedge_{i=0}^{n-1}\foUnifPb_i$ gives a partial substitution $\rho$, such
that $\rho\foUnifPb_n$ falls again in \llambda.
%
% make test ONLY=7003 TEX=TEX + deactivate beta tex
\printAlll
  {{{X,\lambda x.Y},
    {(X\appsep a),a}}}
  {{{A,\lambda x.B},
    {(A\appsep a),a}}}
  {{{Y,B,0},
    {X,A,0}}}
  {{}}

In the example above, we see that $\foUnifPb_1$ instantiates $X$ so that
$\foUnifPb_2$ can be solved in \llambda.
% \noteE[inline]{it is even a ground term, there is no unification left to perform actually}
% \noteD[inline]{i don't understand the note}
On the other hand, we see that, 
\Ue can't solve the compiled problems \hoUnifPb. In
fact, the resolution of $\hoUnifPb_1$ gives the substitution $\sigma = \{ A
\mapsto \lambda x. B\}$, but the dereferencing of $\hoUnifPb_2$ gives the 
non-unifiable problem $(\lambda x. B) \appsep a \nUe a$.

To address
this unification problem, term compilation should capture the
terms in \notllambda and replace them with fresh variables.
This replacement should produce links that we call \linkbeta.

@@

\newcommand{\rhsBetaHead}{\ensuremath{X_{s_1\dots s_n}}}
\newcommand{\rhsBeta}{\ensuremath{\rhsBetaHead\appsep t_1\dots t_m}\xspace}
% \noteE[inline]{Io direi $X_{s_1\ldots s_n}\appsep l_1\ldots l_m$
% ... where S = [s1 .. sn] e L = [l1 .. lm]}

\begin{invariant}[\linkbeta \rhs]
  The \rhs of any \linkbeta has the shape \rhsBeta such that $X$ is a
  unification variable with scope $s_1\dots s_n$ and $t_1\dots t_m$ is a list of
  terms. This is equivalent to \\\elpiIn{app[uva X S | L]}, where \elpiIn{S}~$=
  s_1\dots s_n$ and \elpiIn{L}~$=t_1\dots t_m$.
  \label{inv:beta-rhs}
\end{invariant}

@@

\begin{proof}[Proof sketch]
  Given $X \appsep t_1\dots t_n \Ue t$ where $t$ is a rigid term and
  $t_1\dots t_n$ is not in \llambda. By construction, $X \appsep t_1\dots t_n$
  is replaced with a variable $Y$, and the \linkbeta \linkbetaM{\Gamma}{Y}{X
  \appsep t_1\dots t_n} is created. If $Y$ is instantiated to the rigid term $t$,
  and \rhs is still in \notllambda, then the original \Fo{} unification problem
  is between two terms in \notllambda. Unification fails.
\end{proof}

@@

Detection of \notllambda is quite simple to implement in the compiler, since it
is sufficient to detect applications with flexible head and argument that
are not in \llambda. The following rule for \notllambda compilation is inserted 
just before rule~\ref{rule:compapp}.

@@

% \noteD{Paragraph for decompile in case of \maybebeta}
Decompilation of \linkbeta is possible by extending \elpiIn{commit-link} with
new heuristics.
